Aspects of Dealing with Imperfect Data in Temporal Databases

### 2012-03-16 book/sources/prelim.tex
update
update from the task list.

The most convenient representation for the membership function of a fuzzy number is a triangular function. The membership function $\mu_M$ for the fuzzy set $M$ has also the properties of convexity and normalization. Three values represent a triangular function (see Figure~\ref{fig:triangular}):
\begin{itemize}
\item
$D$ is the singleton value in the core of $\mu_M(x)$.
\item
$D-a$ is the lower value in the support of $\mu_M(x)$. 
\item
$D+b$ is the upper value in the support of $\mu_M(x)$.
\end{itemize}
Note that the values $a$ and $b$ are values in the underlying ordered domain. E.g. $(a,b) \in \mathbb{R}^2$. The notation for a triangular function adopted here is $[D,a,b]$.
\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.4]{graphs/triangular.eps}
  \caption{Triangular function.}
  \label{fig:triangular}
\end{figure}

@@

The general time model represents time as a set with an imposed partial order. There are two main models of time: linear and cyclic. In the linear model, time advances from the past to the future in a totally ordered fashion (a relationship of total order is imposed to the set). A cyclic model of time is used for recurrent processes. The present proposal focuses on the linear model.
\begin{itemize}
\item
Discrete model:  The discrete model of time is isomorphic in relation to the natural numbers. According to this model, each natural number corresponds to an indivisible time unit. The said time unit is called a chronon or, in other words, a 'time quantum', 'moment' or 'instant'. A chronon is the smallest time unit that can be represented by the model.
\item
Dense model: This model is isomorphic in relation to the rational or the real numbers. It represents one instant of time in the gap between another two instants.
\item
Continuous model: This is an extension of the dense model with no gaps. It is isomorphic in relation to the real numbers.
\end{itemize}
Restrictions on time range may be performed on the base of the following two concepts:
\begin{itemize}
\item
Time boundedness: Time can be bounded orthogonally in the past and in the future.
\item
Time distance: As a metric, time has a distance function which presents four properties:
\begin{enumerate}
\item
Distance is non-negative.
\item
The distance between two different elements is not equal to zero.
\item
The distance from $\alpha$ to $\beta$ is the same as that between $\beta$ and $\alpha$.
\item
Triangle inequality: The distance from $\alpha$ to $\gamma$ is equal or greater than the distance from $\alpha$ to $\beta$ plus the distance from $\beta$ to $\gamma$.
\end{enumerate}
\end{itemize}
On the other hand, time may be relative or absolute (in other words, anchored or unanchored). Sometimes, what one considers 'absolute time' is not as definite as one would hope since our concept of absolute time is based on another time reference (for example, January 1st of year one). Relative time has a direction, differently from distance. It is also possible to use negative references so as to represent relative time (e.g. -3 days = three days ago).

@@

A timestamp over $T^p$ is either a time point or a time interval of $T^p$. Time intervals can be represented as sets of time points which may, in turn, be represented by a time interval with the same values at both of its ends. Sometimes a coalesce operator is defined so as to switch between a set of time points and a time interval. Logically, defining a un-coalescing operator is quite complex. 
\subsubsection{\label{subsubsec:timeDb}Time in databases}
We can classify the time recorded in the database through its semantics. As a result, two kinds of time derive from two orthogonal time domains: valid time and transaction time. Valid-time captures the time-varying nature of the reality model. Transaction time structures the update activity associated with the object. The resulting temporal relations are those emerging from the bitemporal conceptual data model BCDM, on which TSQL2 is based.

The domain of valid time is given as $D_{VT}=\lbrace{c_1^v,c_2^v,...,c_k^v}\rbrace$, whereas the domain of transaction time may be given by $D_{TT}=\lbrace{c_1^t,c_2^t,...,c_k^t}\rbrace$. Thus, a valid-time chronon $c^v$  is a member of $D_{VT}$, a transaction-time chronon $c^t$ is a member of $D_{TT}$ and a bitemporal chronon $c^b=(c^t,c^v)$ is an ordered pair of a transaction-time chronon and valid-time chronon.

@@

We say that a fact is current in the relation during each transaction-time chronon when a subset of the transaction-time chronon domain is associated with the valid-time chronon. Only the subset of transaction times less than UC may be associated with a valid time.

While the definition of a bitemporal chronon is symmetric, the relation between valid time and transaction time is asymmetric. It should also be noted that since no two tuples with identical attribute values are allowed in a bitemporal relation instance, the full history of a fact is contained in a single tuple. In graphical representations of bitemporal space, the x-axis represents the transaction-time dimension and the y-axis denotes the valid-time dimension.

There are two special cases of bitemporal relations:
\begin{itemize}
\item
Valid-time relations, which support only valid time.
\item
Transaction-time relations,which support only transaction time.
\end{itemize}
We use the term 'snapshot relation' for a conventional relation. This relation supports neither transaction nor valid time.
\subsubsection{Implicit versus explicit timestamps}
\label{sec:implicitversusexplicittimestamps}
The association between time and facts may be carried out in two ways: explicitly, where the relation is represented by fully explicit timestamps, and implicitly. This distinction is relevant to three aspects of the data model: the update language, the display of data and the query language.

	If the time model works with transaction time, it means that the update language contains implicit timestamps. The transaction-time chronons of the facts are supplied by the system; in the valid-time model, these chronons are supplied by the user. Thus, in a bitemporal data model, the chronons are also supplied by the user.

@@

\subsubsection{\label{subsubsec:models}Temporal data models}
There are several extensions to the relational model that include the temporal element. In the current section, we briefly describe the model and some of its most interesting features. A more detailed summary can be found in \ref{sec:timeindatabasedesign}. Some models work only with valid or transaction time, whereas a few models can handle both. The current study starts with the models supporting valid time only, continues through the models which support transaction time and finally deals with the bitemporal models.

\paragraph{Valid-time models}
Most models support valid time only.

@@

Jones LEGOL 2.0: \cite{legol20} A language used in legislative writing. The temporal order of elements and the valid times for objects are important. It was the first time-oriented algebra to be defined, and some of its features are found in subsequent algebras.
Tuples in this model have two time attributes: Start and Stop. Each one indicates the start point and the end point of the interval where the tuple is valid.

@@

Ariav: \cite{Ariav:1986:TOD:7239.7350} Temporally-oriented data model: a valid-time relation is a set of snapshot relation states indexed by valid time. A calculus-based query language is associated with the model: TOSQL.

@@

Sarda: \cite{Sarda:1990:ESH:627277.627409} This model was designed to support the calculus-based model HSQL. The model associates valid-time with tuples. Objects can either be snapshot or valid-time relations. In this model, valid time is considered to be closed on the left part of the interval and open on the right. Time is represented by an implicit attribute called 'Period' as a single non-atomic value.

@@

Gadia-1:  \cite{Gadia:1988:HRM:49346.50065},\cite{Gadia:1985:QLH:325405.325412} This homogeneous model allows two types of objects: valid-time elements and valid-time relations. The model requires that all attribute values in a given tuple be functions on the same valid-time element. Unlike intervals, valid-time elements are closed under union, difference, and complementation.
Bhargava presents an extension of this model to both valid and transaction time.

Gadia-2: \cite{Gadia:1986:TMM:645471.655410},\cite{Chuen},\cite{Gadia:1988:GMR:971701.50233} Is an extension of the homogeneous model known as multihomogeneous model. Temporal elements may be multi-dimensional to model both valid and transaction time. Attribute values are functions from temporal elements onto attribute value domains, but attribute values do not need to be on the same temporal element. The key attributes in the relations must be single-valued in respect to the interval of validity.

@@

\paragraph{Bitemporal data models}
The bitemporal data models support both valid and transaction time. This section deals with the following seven models of a bitemporal data model.

@@

Ahn: \cite{Snodgrass:1985:TTD:318898.318921},\cite{sno86} This model follows a four-dimensional data model. Each instance is represented as a sequence where each value is composed by a timestamp with a transaction-time chronon and a three dimensional volumes -one of these dimensions being the valid time.

Snodgrass:  \cite{sno86},\cite{Snodgrass:1984:TQL:588011.588041} The data model has four implicit attributes: 
The transaction time when the tuple was inserted.
The transaction time when the tuple was deleted.
The starting point of the valid-time interval.
The end point of the valid time-interval.

@@

McKenzie: \cite{Snodgrass:1993:ATQ:642875.642890},\cite{Mckenzie:1988:ALQ:915060}, \cite{mck81} In this model, the attribute's values are timestamped, but it is required that attributes be single-valued. This makes it possible to perform a Cartesian product. There are two types of objects in the model: the snapshot relations and the valid-time relations. Thus, transaction-time relations are modeled as sequences of snapshot relations. Bitemporal relations are sequences of valid-time relations indexed by transaction time. An attribute's value in a valid-time relation is an ordered pair: a value from the attribute's domain and a set of chronons. Relations are not allowed to have value-equivalent tuples. The model also allows the schema and even the class of the relation (snapshot, transaction-time, valid-time and bitemporal) to change.

@@

Thompson: \cite{171833} This model, based on an accounting system, is called Accounting Data Model. It has four timestamp attributes: two for transaction time, called 'engineering time', and two for valid time, called 'accounting time'. Also, a boolean value called 'timewarp' enables the attribute to change historically.

>>>>

The most convenient representation for the membership function of a fuzzy number is a triangular function. The membership function $\mu_M$ for the fuzzy set $M$ has also the properties of convexity and normalization. Three values represent a triangular function (see Figure~\ref{fig:triangular}):
\begin{itemize}
\item
$D$ is the singleton value in the core of $\mu_M(x)$.
\item
$D-a$ is the lower value in the support of $\mu_M(x)$. 
\item
$D+b$ is the upper value in the support of $\mu_M(x)$.
\end{itemize}
Note that the values $a$ and $b$ are values in the underlying ordered domain. e.g., $(a,b) \in \mathbb{R}^2$. The notation for a triangular function adopted here is $[D,a,b]$.
\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.4]{graphs/triangular.eps}
  \caption{Triangular function.}
  \label{fig:triangular}
\end{figure}

@@

The general time model represents time as a set with an imposed partial order. There are two main models of time: linear and cyclic. In the linear model, time advances from the past to the future in a totally ordered fashion (a relationship of total order is imposed to the set). A cyclic model of time is used for recurrent processes. The present proposal focuses on the linear model.
\begin{itemize}
\item
Discrete model:  The discrete model of time is isomorphic in relation to the natural numbers. According to this model, each natural number corresponds to an indivisible time unit. The said time unit is called a chronon or, in other words, a 'time quantum', 'moment' or 'instant'. A chronon is the smallest time unit that can be represented by the model.
\item
Dense model: This model is isomorphic in relation to the rational or the real numbers. It represents one instant of time in the gap between another two instants.
\item
Continuous model: This is an extension of the dense model with no gaps. It is isomorphic in relation to the real numbers.
\end{itemize}
Restrictions on time range may be performed on the base of the following two concepts:
\begin{itemize}
\item
Time boundedness: Time can be bounded orthogonally in the past and in the future.
\item
Time distance: As a metric, time has a distance function which presents four properties:
\begin{enumerate}
\item
Distance is non-negative.
\item
The distance between two different elements is not equal to zero.
\item
The distance from $\alpha$ to $\beta$ is the same as that between $\beta$ and $\alpha$.
\item
Triangle inequality: The distance from $\alpha$ to $\gamma$ is equal or greater than the distance from $\alpha$ to $\beta$ plus the distance from $\beta$ to $\gamma$.
\end{enumerate}
\end{itemize}
On the other hand, time may be relative or absolute (in other words, anchored or unanchored). Sometimes, what one considers 'absolute time' is not as definite as one would hope since our concept of absolute time is based on another time reference (for example, January 1st of year one). Relative time has a direction, differently from distance. It is also possible to use negative references so as to represent relative time (e.g., -3 days = three days ago).

@@

A timestamp over $T^p$ is either a time point or a time interval of $T^p$. Time intervals can be represented as sets of time points which may, in turn, be represented by a time interval with the same values at both of its ends. Sometimes a coalesce operator is defined so as to switch between a set of time points and a time interval. Logically, defining a un-coalescing operator is quite complex. 
\subsubsection{\label{subsubsec:timeDb}Time in databases}
We can classify the time recorded in the database through its semantics. As a result, two kinds of time derive from two orthogonal time domains: valid-time and transaction-time. Valid-time captures the time-varying nature of the reality model. Transaction time structures the update activity associated with the object. The resulting temporal relations are those emerging from the bitemporal conceptual data model BCDM, on which TSQL2 is based.

The domain of valid-time is given as $D_{VT}=\lbrace{c_1^v,c_2^v,...,c_k^v}\rbrace$, whereas the domain of transaction-time may be given by $D_{TT}=\lbrace{c_1^t,c_2^t,...,c_k^t}\rbrace$. Thus, a valid-time chronon $c^v$  is a member of $D_{VT}$, a transaction-time chronon $c^t$ is a member of $D_{TT}$ and a bitemporal chronon $c^b=(c^t,c^v)$ is an ordered pair of a transaction-time chronon and valid-time chronon.

@@

We say that a fact is current in the relation during each transaction-time chronon when a subset of the transaction-time chronon domain is associated with the valid-time chronon. Only the subset of transaction-times less than UC may be associated with a valid-time.

While the definition of a bitemporal chronon is symmetric, the relation between valid-time and transaction-time is asymmetric. It should also be noted that since no two tuples with identical attribute values are allowed in a bitemporal relation instance, the full history of a fact is contained in a single tuple. In graphical representations of bitemporal space, the x-axis represents the transaction-time dimension and the y-axis denotes the valid-time dimension.

There are two special cases of bitemporal relations:
\begin{itemize}
\item
Valid-time relations, which support only valid-time.
\item
Transaction-time relations,which support only transaction-time.
\end{itemize}
We use the term 'snapshot relation' for a conventional relation. This relation supports neither transaction nor valid-time.
\subsubsection{Implicit versus explicit timestamps}
\label{sec:implicitversusexplicittimestamps}
The association between time and facts may be carried out in two ways: explicitly, where the relation is represented by fully explicit timestamps, and implicitly. This distinction is relevant to three aspects of the data model: the update language, the display of data and the query language.

	If the time model works with transaction-time, it means that the update language contains implicit timestamps. The transaction-time chronons of the facts are supplied by the system; in the valid-time model, these chronons are supplied by the user. Thus, in a bitemporal data model, the chronons are also supplied by the user.

@@

\subsubsection{\label{subsubsec:models}Temporal data models}
There are several extensions to the relational model that include the temporal element. In the current section, we briefly describe the model and some of its most interesting features. A more detailed summary can be found in \ref{sec:timeindatabasedesign}. Some models work only with valid or transaction-time, whereas a few models can handle both. The current study starts with the models supporting valid-time only, continues through the models which support transaction-time and finally deals with the bitemporal models.

\paragraph{Valid-time models}
Most models support valid-time only.

@@

Jones LEGOL 2.0: \cite{legol20} A language used in legislative writing. The temporal order of elements and the valid-times for objects are important. It was the first time-oriented algebra to be defined, and some of its features are found in subsequent algebras.
Tuples in this model have two time attributes: Start and Stop. Each one indicates the start point and the end point of the interval where the tuple is valid.

@@

Ariav: \cite{Ariav:1986:TOD:7239.7350} Temporally-oriented data model: a valid-time relation is a set of snapshot relation states indexed by valid-time. A calculus-based query language is associated with the model: TOSQL.

@@

Sarda: \cite{Sarda:1990:ESH:627277.627409} This model was designed to support the calculus-based model HSQL. The model associates valid-time with tuples. Objects can either be snapshot or valid-time relations. In this model, valid-time is considered to be closed on the left part of the interval and open on the right. Time is represented by an implicit attribute called 'Period' as a single non-atomic value.

@@

Gadia-1:  \cite{Gadia:1988:HRM:49346.50065},\cite{Gadia:1985:QLH:325405.325412} This homogeneous model allows two types of objects: valid-time elements and valid-time relations. The model requires that all attribute values in a given tuple be functions on the same valid-time element. Unlike intervals, valid-time elements are closed under union, difference, and complementation.
Bhargava presents an extension of this model to both valid and transaction-time.

Gadia-2: \cite{Gadia:1986:TMM:645471.655410},\cite{Chuen},\cite{Gadia:1988:GMR:971701.50233} Is an extension of the homogeneous model known as multihomogeneous model. Temporal elements may be multi-dimensional to model both valid and transaction-time. Attribute values are functions from temporal elements onto attribute value domains, but attribute values do not need to be on the same temporal element. The key attributes in the relations must be single-valued in respect to the interval of validity.

@@

\paragraph{Bitemporal data models}
The bitemporal data models support both valid and transaction-time. This section deals with the following seven models of a bitemporal data model.

@@

Ahn: \cite{Snodgrass:1985:TTD:318898.318921},\cite{sno86} This model follows a four-dimensional data model. Each instance is represented as a sequence where each value is composed by a timestamp with a transaction-time chronon and a three dimensional volumes -one of these dimensions being the valid-time.

Snodgrass:  \cite{sno86},\cite{Snodgrass:1984:TQL:588011.588041} The data model has four implicit attributes: 
The transaction-time when the tuple was inserted.
The transaction-time when the tuple was deleted.
The starting point of the valid-time interval.
The end point of the valid-time-interval.

@@

McKenzie: \cite{Snodgrass:1993:ATQ:642875.642890},\cite{Mckenzie:1988:ALQ:915060}, \cite{mck81} In this model, the attribute's values are timestamped, but it is required that attributes be single-valued. This makes it possible to perform a Cartesian product. There are two types of objects in the model: the snapshot relations and the valid-time relations. Thus, transaction-time relations are modeled as sequences of snapshot relations. Bitemporal relations are sequences of valid-time relations indexed by transaction-time. An attribute's value in a valid-time relation is an ordered pair: a value from the attribute's domain and a set of chronons. Relations are not allowed to have value-equivalent tuples. The model also allows the schema and even the class of the relation (snapshot, transaction-time, valid-time and bitemporal) to change.

@@

Thompson: \cite{171833} This model, based on an accounting system, is called Accounting Data Model. It has four timestamp attributes: two for transaction-time, called 'engineering time', and two for valid-time, called 'accounting time'. Also, a boolean value called 'timewarp' enables the attribute to change historically.

### 2012-03-16 book/sources/proposal.tex
update from the task list.

% Proposal
%
%
The model deals with uncertainty in the valid time. A valid time period  \emph{VT} is specified with two points: a starting time point \emph{St} and an ending time point \emph{Et}. Thus, a fuzzy validity period, \emph{FVP} is a valid time period in which the starting, the ending or both points are not precisely known. More formally we make the following definitions:

>>>>

% Proposal
%
%
The model deals with uncertainty in the valid-time. A valid-time period  \emph{VT} is specified with two points: a starting time point \emph{St} and an ending time point \emph{Et}. Thus, a fuzzy validity period, \emph{FVP} is a valid-time period in which the starting, the ending or both points are not precisely known. More formally we make the following definitions:
